---
title: "Advanced Usage and Technical Details"
author: "Max Rubinstein, Denis Agniel, Larry Han, Marcela Horvitz-Lennon, Sharon-Lise Normand"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Usage and Technical Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette covers advanced usage of **jointbounds**, including:
- Custom SuperLearner configurations
- Understanding nuisance parameter estimation
- Working with influence functions
- Technical options and performance tuning

```{r}
library(jointbounds)
suppressPackageStartupMessages(library(SuperLearner))
```

## SuperLearner Configuration

### Available Libraries

SuperLearner offers many prediction algorithms. Common choices:

```{r, eval = FALSE}
# See all available algorithms
listWrappers()
```

**Regression algorithms:**
- `SL.glm` - Logistic/linear regression (fast, parametric)
- `SL.gam` - Generalized additive models (flexible, smooth)
- `SL.ranger` - Random forests (flexible, nonparametric)
- `SL.xgboost` - Gradient boosting (powerful, tunable)
- `SL.earth` - MARS (multivariate adaptive regression splines)
- `SL.glmnet` - Elastic net (regularized regression)

**Screening algorithms:**
- `screen.corP` - Select by correlation p-value
- `screen.glmnet` - LASSO screening

### Custom Library Specifications

You can create ensembles by providing vectors of algorithms:

```{r}
# Example data
set.seed(20260313)
n <- 500
X <- runif(n, -2, 2)
A <- rbinom(n, 1, plogis(X))
C <- rbinom(n, 1, 0.2 + 0.1 * A)
Y <- ifelse(C == 1, NA, rbinom(n, 1, plogis(0.5 * A + 0.3 * X)))
dat <- data.frame(Y = Y, A = A, C = C, X = X)

# Simple ensemble
fit_ensemble <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "ate",
  assumption = "general",
  sl_lib = c("SL.glm", "SL.gam", "SL.mean"),
  V = 2,
  seed = 1
)

# SuperLearner automatically selects weights
# (Weights not directly accessible in output)
```

### Separate Libraries for Each Nuisance Function

For maximum flexibility, specify different libraries:

```{r, eval = FALSE}
fit_custom <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "ate",
  assumption = "general",
  sl_lib_prop = c("SL.glm", "SL.gam"),           # Propensity score
  sl_lib_miss = c("SL.glm", "SL.ranger"),        # Missingness mechanism
  sl_lib_outcome = c("SL.glm", "SL.xgboost"),    # Outcome regression
  V = 5,  # More folds for stable SuperLearner
  seed = 1
)
```

### When to Use Different Libraries

**Simple parametric (`SL.glm` only):**
- Small samples (n < 500)
- When you want fast computation
- When relationships are approximately linear
- For debugging/prototyping

**Flexible ensembles:**
- Larger samples (n > 1000)
- Unknown functional forms
- Complex covariate relationships
- Final analyses

## Cross-Fitting and Sample Splitting

### Understanding V-Fold Cross-Fitting

The package uses cross-fitting to estimate nuisance functions. Parameter `V` controls the number of folds:

```{r}
# V = 2 (default): Split data in half
fit_v2 <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  estimand = "ate", assumption = "general",
  V = 2, sl_lib = "SL.glm", seed = 1
)

# V = 5: More folds (better for larger samples)
fit_v5 <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  estimand = "ate", assumption = "general",
  V = 5, sl_lib = "SL.glm", seed = 1
)

# V = 1: No cross-fitting (not recommended for inference)
fit_v1 <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  estimand = "ate", assumption = "general",
  V = 1, sl_lib = "SL.glm", seed = 1
)
```

**Guidelines:**
- V = 2 is sufficient for most applications
- V = 5-10 for very large samples or complex SuperLearner libraries
- V = 1 only when using simple parametric models (SL.glm only)

### Accessing Fold Information

```{r}
# Check fold assignments
table(fit_v2$nuisance$fold_id)

# Nuisance estimates are out-of-sample within each fold
head(fit_v2$nuisance$fold_id)
```

## Working with Influence Functions

### Extracting Influence Functions

The influence function matrix φ contains information used for inference:

```{r}
phi <- fit_v2$phi
dim(phi)  # n x 6 matrix
colnames(phi)
```

The 6 columns correspond to:
1. `phi_psi_naive`: Naive estimator (complete-case)
2-3. Columns for lower bound computation
4-5. Columns for upper bound computation
6. Additional component

### Custom Inference

You can use the influence functions for custom inference:

```{r}
# Standard error calculation
se_custom <- function(phi_col) {
  sqrt(var(phi_col) / length(phi_col))
}

# Custom confidence interval
estimate <- fit_v2$result$lower
se <- fit_v2$result$se_lower
ci_lower <- estimate - 1.96 * se
ci_upper <- estimate + 1.96 * se

cat("Custom 95% CI for lower bound: [", ci_lower, ",", ci_upper, "]\n")
```

### Bootstrap with Custom Parameters

Use influence functions to compute bounds at new parameter values without re-estimating nuisances:

```{r, eval = FALSE}
# Extract fitted object components
phi <- fit_v2$phi
nuisance <- fit_v2$nuisance

# Use multiplier_bootstrap_grid for custom parameter values
new_bounds <- multiplier_bootstrap_grid(
  fit_v2,
  param_grid = data.frame(delta_0u = 0.6, delta_1u = 0.6),
  bound_spec = "lower",
  assumption = "bounded_delta",
  B = 100
)
```

## Outcome Model Stratification

### Stratified vs. Pooled Outcome Models

By default, outcome models are fit separately within treatment arms:

```{r}
# Stratified (default): E[Y|A=a,X] fit separately for a=0 and a=1
fit_stratified <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  estimand = "ate", assumption = "general",
  stratify_mu = TRUE,
  V = 2, sl_lib = "SL.glm", seed = 1
)

# Pooled: E[Y|A,X] fit once with A as covariate
fit_pooled <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  estimand = "ate", assumption = "general",
  stratify_mu = FALSE,
  V = 2, sl_lib = "SL.glm", seed = 1
)
```

**When to use pooled:**
- Small observed samples within treatment arms
- Believe treatment effect is approximately constant
- Want to borrow strength across treatment groups

## Assumption Types in Detail

### General Bounds

No assumptions beyond conditional independence of missingness given (A, X):

```{r}
fit_general <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  estimand = "ate",
  assumption = "general",
  V = 2, sl_lib = "SL.glm", seed = 1
)
```

### Bounded Delta (Informative Proportion)

At most δ proportion of missingness is informative:

```{r}
fit_bounded_delta <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  estimand = "psi1",
  assumption = "bounded_delta",
  delta_0u = 0.8,  # Upper bound on proportion
  delta_1u = 0.8,
  V = 2, sl_lib = "SL.glm", seed = 1
)
```

### Monotonicity

Missingness is more likely for one outcome value:

```{r}
# Positive monotonicity: P(C=1|Y=1,A,X) >= P(C=1|Y=0,A,X)
fit_mono_pos <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  estimand = "ate",
  assumption = "monotonicity_pos",
  delta_0u = 0.8,
  delta_1u = 0.8,
  V = 2, sl_lib = "SL.glm", seed = 1
)

# Negative monotonicity: reverse inequality
fit_mono_neg <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  estimand = "ate",
  assumption = "monotonicity_neg",
  delta_0u = 0.8,
  delta_1u = 0.8,
  V = 2, sl_lib = "SL.glm", seed = 1
)
```

### Bounded Risk

Bounds on the ratio of missing-to-observed outcome risks:

```{r}
fit_bounded_risk <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  estimand = "ate",
  assumption = "bounded_risk",
  delta_0u = 0.8,
  delta_1u = 0.8,
  tau_0 = 2,  # Risk ratio bound in control
  tau_1 = 2,  # Risk ratio bound in treatment
  V = 2, sl_lib = "SL.glm", seed = 1
)
```

### Point Identification

When sensitivity parameters are known:

```{r}
fit_point <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  estimand = "ate",
  assumption = "point_ate",
  delta_0 = 0.5,  # Known value
  delta_1 = 0.5,
  tau = 2,
  V = 2, sl_lib = "SL.glm", seed = 1
)
```

## Multiplier Bootstrap Options

### Multiplier Types

Two types of multipliers are available:

```{r, eval = FALSE}
# Rademacher (default): Random ±1
fit_rademacher <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  param_grid = list(delta_0 = c(0.5, 0.7, 0.9),
                    delta_1 = c(0.5, 0.7, 0.9)),
  multiplier = "rademacher",
  B = 500, seed = 1
)

# Gaussian: Random N(0,1)
fit_gaussian <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  param_grid = list(delta_0 = c(0.5, 0.7, 0.9),
                    delta_1 = c(0.5, 0.7, 0.9)),
  multiplier = "gaussian",
  B = 500, seed = 1
)
```

Both are valid; results should be similar. Rademacher is slightly faster.

### Number of Bootstrap Replicates

```{r, eval = FALSE}
# Quick exploration
fit_quick <- mar_bounds(..., B = 100)

# Standard analysis
fit_standard <- mar_bounds(..., B = 500)

# Publication-quality
fit_publication <- mar_bounds(..., B = 1000)
```

## Performance Optimization

### Small Samples (n < 200)

```{r, eval = FALSE}
fit_small <- mar_bounds(
  dat,
  Y = "Y", A = "A", C = "C", X = "X",
  sl_lib = "SL.glm",  # Parametric only
  V = 1,              # No cross-fitting
  B = 200,            # Fewer bootstrap replicates
  seed = 1
)
```

### Large Samples (n > 5000)

```{r, eval = FALSE}
fit_large <- mar_bounds(
  dat,
  Y = "Y", A = "A", C = "C", X = "X",
  sl_lib = c("SL.glm", "SL.gam", "SL.ranger"),  # Can use complex models
  V = 5,              # More folds for stability
  B = 1000,           # More bootstrap replicates
  seed = 1
)
```

### Parameter Grid Size

For very large grids, consider computing only necessary bounds:

```{r, eval = FALSE}
# Only compute one bound
fit_lower <- mar_bounds(..., grid_bounds = "lower")
fit_upper <- mar_bounds(..., grid_bounds = "upper")

# Or split into smaller grids
grid1 <- list(delta_0 = seq(0.3, 0.6, 0.1), delta_1 = seq(0.3, 0.6, 0.1))
grid2 <- list(delta_0 = seq(0.7, 1.0, 0.1), delta_1 = seq(0.7, 1.0, 0.1))
```

## Troubleshooting

### SuperLearner Convergence Issues

If you see "All algorithms dropped from library":

```{r, eval = FALSE}
# Solution 1: Use simpler library
mar_bounds(..., sl_lib = "SL.glm")

# Solution 2: Increase V to avoid small training samples
mar_bounds(..., V = 2)  # instead of V = 5

# Solution 3: Add fallback algorithms
mar_bounds(..., sl_lib = c("SL.glm", "SL.mean"))
```

### Missing Value Warnings

The package expects missingness to be coded with `NA` in the outcome and `C = 1`:

```{r, eval = FALSE}
# Correct coding
dat$Y[dat$C == 1] <- NA

# Check consistency
stopifnot(all(is.na(dat$Y) == (dat$C == 1)))
```

### Reproducibility

Always set a seed for reproducible results:

```{r, eval = FALSE}
fit <- mar_bounds(..., seed = 20260313)  # YYYYMMDD format recommended
```

## References

Rubinstein, M., Agniel, D., Han, L., Horvitz-Lennon, M., & Normand, S.-L. (2026).
Bounding causal effects with an unknown mixture of informative and non-informative missingness.
Journal of Causal Inference (Accepted). https://arxiv.org/pdf/2411.16902

van der Laan, M. J., Polley, E. C., & Hubbard, A. E. (2007).
Super learner. Statistical Applications in Genetics and Molecular Biology, 6(1).
