---
title: "Continuous-outcome L2 sensitivity analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Continuous-outcome L2 sensitivity analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE}
library(marbounds)
```

## A moment-valid continuous-outcome simulation

The data generator uses Gaussian potential-outcome errors. Consequently,
the population fourth moments required by the paper's closed-form interaction
bound are finite. Informative type depends on the factual outcome, all
informative-type units are missing, and noninformative missingness is
conditionally independent of the outcome. By default, treatment also depends
on a disturbance shared by both potential outcomes, giving nonzero residual
confounding. Set `confounding_strength = 0` to recover conditional
exchangeability.

The `violation` argument selects which assumptions fail:

```{r dgp-regimes}
d_mar <- simulate_l2_data(500, violation = "mar")
d_nuc <- simulate_l2_data(500, violation = "nuc")
d_both <- simulate_l2_data(500, violation = "both")
rbind(
  mar = unlist(l2_oracle_sensitivity_parameters(d_mar)[c("delta_R", "delta_A")]),
  nuc = unlist(l2_oracle_sensitivity_parameters(d_nuc)[c("delta_R", "delta_A")]),
  both = unlist(l2_oracle_sensitivity_parameters(d_both)[c("delta_R", "delta_A")])
)
```

```{r simulate}
dat <- simulate_l2_data(n = 1000, seed = 2026)
mean(dat$C)
mean(dat$Y1 - dat$Y0)
mean((dat$Yfull - mean(dat$Yfull))^4)
mean(dat$U_I)
mean(dat$U_I)/mean(dat$C)
l2_oracle_sensitivity_parameters(dat)
```

The output also contains both potential outcomes and potential missingness
indicators, as well as true DGP quantities such as `p_A1_given_X`,
`p_UI0_given_YX`, `p_C0_given_YX`, and `p_C0_given_X`. For continuous outcomes,
`mu_Y0`, `sd_Y0`, and `f_Y0_given_X` describe the conditional law of `Y0`.

The last quantity is only an empirical diagnostic. No finite dataset can prove
that a population fourth moment exists; that condition must also be justified
from the outcome's scientific and statistical model.

## Point-identified ATE under MAR and no unmeasured confounding

When both identifying assumptions are maintained, `generate_ate()` returns the
cross-fitted AIPW estimate of the ATE, arm-specific means, Wald intervals,
influence scores, and positivity diagnostics:

```{r identified-ate}
identified_dat <- simulate_l2_data(1000, violation = "none", seed = 2027)
identified <- generate_ate(
  identified_dat, "Y", "A", "C", c("X1", "X2"),
  folds = 5, nuisance_method = "SuperLearner"
)
identified$ate
identified$diagnostic_summary
```

The estimator is consistent if the outcome regressions are correct or if the
arm-specific joint treatment-and-response probabilities are correct. This is
the point-identified baseline to which the sensitivity intervals can be
compared.

## Closed-form outer bounds

```{r cs}
cs <- l2_bounds(
  dat, Y = "Y", A = "A", C = "C", X = c("X1", "X2"),
  delta_M = c(.20, .20), delta_K = c(.15, .15),
  model = "net", method = "cs", folds = 5,
  nuisance_method = "SuperLearner", return_true_bounds = TRUE
)
cs$ate
cs$true_bounds$ate
```

The propensity, observation probability, and outcome regression are estimated
as functions of `X` with cross-fitting. Use `l2_cs_crossfit()` for the
orthogonal EIF estimator and pointwise Wald inference for the nonlinear CS
moments. The oracle result is available only because the simulated data contain
the known DGP columns.

Here `delta_M` bounds the global L2 net distortion of the factual outcome law
relative to the respondent law, and `delta_K` bounds the global L2 strength of
transport across factual treatment groups. The package accepts `delta_A` as a
backward-compatible alias for `delta_K`.

## Empirical sharp sieve-dual bounds

The sharp computation enforces conditional normalization on the supplied
basis. The default uses an intercept and main effects for `X`.

```{r sharp, eval=FALSE}
sharp <- l2_bounds(
  dat, Y = "Y", A = "A", C = "C", X = c("X1", "X2"),
  delta_M = .20, delta_K = .15, model = "net",
  method = "both", folds = 5,
  basis = ~ X1 + X2 + I(X1^2) + I(X2^2)
)
sharp$ate
sharp$diagnostic_summary
sharp$diagnostic_table
```

The endpoint-level table reports optimizer convergence, the primal--dual gap,
normalization error, achieved-versus-allowed divergence budgets, the two global
multipliers, the fraction of observations on each pointwise boundary, and local
Hessian curvature. The one-row summary says whether any endpoint was flagged.
A fit can converge numerically but still receive a regularity flag because a
budget is slack, a multiplier is nearly zero, or local curvature is weak. The
function issues one consolidated warning rather than a separate warning for
each check. The original nested output remains in `sharp$diagnostics`.

Diagnostic tolerances can be changed with `diagnostic_control`; for example,
`diagnostic_control = list(compute_hessian = FALSE)` skips the additional local
curvature calculation, while `warn = FALSE` suppresses only the warning and not
the returned diagnostics.

### Cross-fitted plug-in and DR inference

`l2_sharp_crossfit()` refits both nuisance regressions and the sieve-dual
optimization on each training fold. On its held-out fold, it reports an
inverse-respondent-weighted plug-in score and a DR score. The latter estimates
the conditional mean of the optimized loss, `m_H(X)`, by regressing the
training-fold respondent pseudo-outcome on `X`; it also includes the analytic
chain-rule corrections for the treatment propensity and response probability.

```{r sharp-dr, eval=FALSE}
cf <- l2_sharp_crossfit(
  dat, "Y", "A", "C", c("X1", "X2"),
  delta_M = .20, delta_K = .15, model = "net",
  folds = 5, nuisance_method = "SuperLearner",
  basis = ~ X1 + X2 + I(X1^2) + I(X2^2)
)
cf$arm
cf$ate
```

The returned Wald intervals describe sampling uncertainty for the empirical
sieve endpoint. They require stable active sets and sufficiently fast nuisance
and dual estimation; active-set changes can make the endpoint nonregular.

The simulation wrapper compares endpoint bias, RMSE, standard-error
calibration, and interval coverage for the two evaluations. Supplying known
endpoint truth is preferable. If it is omitted, the function constructs a
large-sample empirical-sieve reference and labels it as such.

```{r sharp-dr-simulation, eval=FALSE}
mc <- l2_compare_dr_plugin(
  B = 500, n = 1000, truth_n = 50000,
  delta_M = .20, delta_K = .15, model = "net",
  nuisance_simulation = "oracle_error",
  nuisance_error_rate = .25,
  nuisance_error_scale = c(e = 1, rho = 1, mu = 1, regression = 1),
  nuisance_error_sign = c(e = 1, rho = -1, mu = 1, regression = -1)
)
mc$summary
mc$bound_coverage
mc$oracle_parameters
mc$evaluated_parameters
mc$linf_bounds
mc$linf_sharp_bounds
mc$bound_comparison
```

The complete paper design is available as data rather than being hidden in a
script. The high-level runner checkpoints after every scenario and returns
tidy endpoint, confidence-set, width, mechanism, and optimizer summaries.

```{r paper-simulation, eval=FALSE}
design <- l2_simulation_design("fast")
study <- l2_simulation_study(design, B = 10, truth_n = 2000, folds = 3)
study$endpoint_summary
study$set_coverage
plot(study, "coverage")

calibration_mc <- l2_calibration_simulation(
  B = 100, n = 1000, truth_n = 30000, p_z = 0:1
)
calibration_mc$summary
```

Every comparison reports `plugin`, `dr` (the EIF with the specified nuisance
errors), and `true_eif`. The last uses exact DGP treatment and response
probabilities and numerical integration under the known conditional respondent
outcome law. In controlled-error mode the code fits no regressions and therefore
does not cross-fit. It directly perturbs the treatment, response, outcome-
regression, and second-stage conditional-regression functions. For the sharp
bound it also recomputes the dual solution and (H) using the perturbed
nuisances. For the CS bound, the conditional residual moments are perturbed
separately from (mu_a).

The minimal oracle parameters are returned on every run. If no sensitivity
radii are supplied, `parameter_evaluation = "auto"`
evaluates only at those oracle parameters. Set `parameter_evaluation =
"oracle"` to override supplied radii, or `"specified"` to evaluate the supplied
radii while retaining the oracle values as diagnostics.

The default bounded latent link makes the oracle density ratios uniformly
bounded. The output therefore also compares the L2 sharp-sieve interval with a
valid L-infinity/L1 Holder outer interval. This L-infinity interval is an outer
benchmark rather than a sharp conditional-normalization solution. With
`density_ratio_link = "linear"`, the active density ratios are unbounded and
the package reports infinite L-infinity radii and an unbounded interval.

Set `linf_method = "sharp"` or `"both"` to solve the pointwise-envelope problem
as a linear program. This enforces the same empirical basis normalization for
`M` and `L` used by the L2 sieve program, making the comparison sharp for the
chosen empirical sieve. It requires the suggested `lpSolve` package. The
`linf_max_n` argument caps the calibration sample used by the LP because it has
two variables per respondent and pointwise constraints per row.

For an observed dataset rather than the known simulation law, use the
data-facing comparison directly:

```{r linf-data, eval=FALSE}
l2_linf_bounds(
  dat, "Y", "A", "C", c("X1", "X2"), delta = .5,
  delta_R_inf = .4, delta_A_inf = .3, method = "both"
)
```

Here each oracle nuisance is perturbed in a fixed smooth direction with
magnitude proportional to `n^(-nuisance_error_rate)`. Fixing the direction
prevents first-order plug-in bias from disappearing because positive and
negative perturbations average out across replications. A zero rate gives
persistent misspecification. Use `nuisance_simulation = "estimated"` for the
slower experiment that refits the requested learners in every replication.

## Primary net-missingness model

```{r net}
net <- l2_bounds(
  dat, Y = "Y", A = "A", C = "C", X = c("X1", "X2"),
  model = "net", delta_M = c(.12, .12), delta_K = c(.15, .15),
  method = "cs", folds = 5
)
net$ate
```

This model bounds the net distributional distortion `M - 1`. It is the
package default and avoids separately eliciting latent informative-missingness
prevalence and severity. Use `model = "separated"` with `delta`, `delta_R`,
and `delta_K` for the mechanistic comparison.

Both cross-fitted inference functions also accept `model = "net"` and
`delta_M`.

The same interface nests analyses that maintain one identifying assumption.
Set `delta_M = 0` to maintain MAR and vary only `delta_K`; set
`delta_K = 0` to maintain conditional exchangeability and vary only
`delta_M`. When the corresponding affine density-ratio extremizer satisfies
the pointwise restriction and the basis enforces the needed normalization,
the sharp and Cauchy--Schwarz endpoints coincide.

```{r single-mechanism}
nuc_only <- l2_bounds(
  dat, "Y", "A", "C", c("X1", "X2"),
  delta_M = 0, delta_K = .15, model = "net", method = "both"
)
mar_only <- l2_bounds(
  dat, "Y", "A", "C", c("X1", "X2"),
  delta_M = .12, delta_K = 0, model = "net", method = "both"
)
```

## Sensitivity surface and tipping frontier

```{r grid}
grid <- expand.grid(
  delta_M = seq(0, .4, length.out = 5),
  delta_K = seq(0, .3, length.out = 5)
)
surface <- l2_sensitivity_grid(
  dat, "Y", "A", "C", c("X1", "X2"), grid = grid,
  model = "net", method = "cs", folds = 5
)
head(surface$results)
frontier <- l2_tipping_frontier(surface, threshold = 0, endpoint = "lower")
if (any(frontier$tipped_on_grid)) plot(frontier)
```

When `method` includes `"sharp"`, `surface$diagnostic_table` has four rows per
grid pair (two arms by two endpoints), and `surface$diagnostic_summary` has one
row per pair. A consolidated warning lists the affected `grid_id` values, so a
large surface does not produce a stream of endpoint-specific warnings.

Groups for which the null is not reached have `tipped_on_grid = FALSE` rather
than being mislabeled with the closest grid point.

For simultaneous rather than pointwise uncertainty, evaluate the cross-fitted
scores on a common grid and apply the paper's multiplier process:

```{r multiplier-band, eval=FALSE}
band <- l2_sensitivity_band(
  dat, "Y", "A", "C", c("X1", "X2"), grid,
  model = "net", bound_method = "cs", estimator = "eif",
  folds = 5, B = 1000, keep_fits = TRUE
)
band$outward
```

The same grid can be inverted. For an observed benchmark discrepancy, the
calibration frontier reports the smallest confounding radius on the grid that,
together with each missingness radius, can reproduce that discrepancy:

```{r inverse-calibration}
l2_calibration_frontier(
  surface,
  benchmark = 0.1,
  type = "discrepancy"
)
```

This is a frontier of jointly sufficient budgets, not an estimate of the true
decomposition. Because `surface` above uses CS bounds, its inversion is an
outer approximation. A `sharp_sieve` grid yields the corresponding empirical
sieve calibration frontier.

## Calibration and bootstrap uncertainty

When MAR and exchangeability may both fail, mechanism-specific calibration is
entangled. The primary joint calibration therefore uses a reduced-form `M`
tilt, interpreted as total rather than mechanism-specific distortion:

```{r calibration}
l2_calibrate(M = c(.95, 1, 1.05), benchmark = "observed_data")
```

A mechanism-specific observed-data benchmark must state what identifies the
other mechanism. For example, calibrating `K` requires
`conditional_on = "MAR"`; calibrating `R` requires
`conditional_on = "exchangeability"`. The function rejects an observed-data
pair of `R` and `K` benchmarks because that pair is not jointly identified.

The main joint calibration uses paired covariate-omission AIPW benchmarks:

```{r benchmark-calibration, eval=FALSE}
bench <- l2_aipw_benchmarks(
  dat, "Y", "A", "C", c("X1", "X2"), p_z = 0:1,
  estimator = "aipw", variance = TRUE
)
bench_band <- l2_benchmark_band(bench, B = 1000)
frontier_band <- l2_calibration_band(band, bench, B = 1000)
plot(frontier_band)
calibrated <- l2_benchmark_frontiers(surface, bench)
l2_equal_radius_comparison(surface, calibrated, plot = TRUE)
```

A full nonparametric bootstrap refits the nuisance functions and bounds. Sharp
bounds may be nonregular at active-set transitions, so percentile intervals
should be treated cautiously in that regime.

```{r bootstrap, eval=FALSE}
boot <- l2_bootstrap(
  dat, "Y", "A", "C", c("X1", "X2"),
  delta = .5, delta_R = .25, delta_A = .15,
  method = "cs", B = 200
)
boot$ci
```

## Binary composite and separable effects

For binary outcomes, `l2_prevalence_bounds()` implements the MAR-only capped
L2 support-function bounds for the composite and separable direct effects.
The influence score includes the multiplier contribution from the active norm
constraint because the norm is taken under the observed covariate law.

```{r binary-alternative, eval=FALSE}
binary_dat <- dat
binary_dat$Y <- as.numeric(binary_dat$Y > 0)
l2_prevalence_bounds(
  binary_dat, "Y", "A", "C", c("X1", "X2"),
  estimand = "composite", kappa = c(.2, .2), folds = 5
)
```
