---
title: "Introduction to jointbounds"
author: "Max Rubinstein, Denis Agniel, Larry Han, Marcela Horvitz-Lennon, Sharon-Lise Normand"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to jointbounds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The **jointbounds** package implements methods for bounding causal effects when outcomes are missing under a mixture of informative and non-informative missingness. This vignette introduces the basic usage of the package.

### Key Features

- **Generic data interface**: Works with standard `data.frame` objects
- **Flexible assumptions**: From general bounds to point identification
- **Multiple estimands**: ATE, composite effects (Ψ₁), separable direct effects (Ψ₂)
- **SuperLearner integration**: Flexible machine learning for nuisance parameter estimation
- **Inference**: Asymptotic confidence intervals and multiplier bootstrap for parameter grids

## Installation

```{r, eval = FALSE}
# Install the development version from GitHub
remotes::install_github("mrubinst757/jointbounds")
```

Load the package and SuperLearner:

```{r}
library(jointbounds)
suppressPackageStartupMessages(library(SuperLearner))
```

## Basic Example

### Data Structure

The package requires a `data.frame` with:
- `Y`: Outcome variable (may contain `NA` for missing values)
- `A`: Treatment variable (binary: 0/1)
- `C`: Missingness indicator (1 = missing, 0 = observed)
- `X`: Covariates (can be multiple columns)

### Simulated Data

Let's create a simple dataset:

```{r}
set.seed(20260313)
n <- 500

# Covariates
X <- runif(n, -2, 2)

# Treatment (depends on X)
A <- rbinom(n, 1, plogis(X))

# Missingness (depends on treatment)
C <- rbinom(n, 1, 0.2 + 0.1 * A)

# Outcome (only observed when C = 0)
Y <- ifelse(C == 1, NA, rbinom(n, 1, plogis(0.5 * A + 0.3 * X)))

# Create data frame
dat <- data.frame(Y = Y, A = A, C = C, X = X)

# Summary
summary(dat)
cat("\nMissing outcomes:", sum(is.na(dat$Y)), "/", n, "\n")
```

## Computing Bounds

### General Bounds (No Assumptions)

The most conservative approach makes no assumptions about the missingness mechanism:

```{r}
fit_general <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "ate",
  assumption = "general",
  sl_lib = "SL.glm",
  V = 2,
  seed = 1
)

# View results
fit_general$result
```

The output shows:
- **naive**: Complete-case estimate (assumes MCAR)
- **lower/upper**: Bounds on the ATE
- **ci_lower_*/ci_upper_***: Confidence intervals for each bound

### Bounded Proportion of Informative Missingness

If we're willing to assume that at most a proportion δ of the missingness is informative, we can tighten the bounds:

```{r}
fit_bounded_delta <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "ate",
  assumption = "bounded_delta",
  delta_0u = 0.8,  # At most 80% informative in control group
  delta_1u = 0.8,  # At most 80% informative in treatment group
  sl_lib = "SL.glm",
  V = 2,
  seed = 1
)

fit_bounded_delta$result
```

Notice the bounds are tighter than the general bounds.

### Point Identification

If we know the exact sensitivity parameters (e.g., from a validation study), we can get point identification:

```{r}
fit_point <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "ate",
  assumption = "point_ate",
  delta_0 = 0.5,
  delta_1 = 0.5,
  tau = 2,
  sl_lib = "SL.glm",
  V = 2,
  seed = 1
)

fit_point$result
```

## Understanding the Output

The main output is a `result` data frame with:

**For point estimates:**
- `estimate`: Point estimate
- `se`: Standard error
- `ci_lower`, `ci_upper`: Confidence interval

**For bounds:**
- `naive`: Complete-case estimate
- `lower`, `upper`: Lower and upper bound estimates
- `se_lower`, `se_upper`: Standard errors for bounds
- `ci_lower_lower`, `ci_lower_upper`: CI for lower bound
- `ci_upper_lower`, `ci_upper_upper`: CI for upper bound

## Comparing Assumptions

Let's compare different assumptions:

```{r}
# Add delta=0.5 bounds
fit_delta_05 <- mar_bounds(
  dat, Y = "Y", A = "A", C = "C", X = "X",
  estimand = "ate", assumption = "bounded_delta",
  delta_0u = 0.5, delta_1u = 0.5,
  sl_lib = "SL.glm", V = 2, seed = 1
)

results <- data.frame(
  assumption = c("General", "Delta=0.8", "Delta=0.5", "Point ID"),
  lower = c(
    fit_general$result$lower,
    fit_bounded_delta$result$lower,
    fit_delta_05$result$lower,
    NA   # Point estimate doesn't have bounds
  ),
  upper = c(
    fit_general$result$upper,
    fit_bounded_delta$result$upper,
    fit_delta_05$result$upper,
    NA
  ),
  point = c(NA, NA, NA, fit_point$result$estimate)
)

print(results)
```

As we make stronger assumptions (smaller δ), the bounds get tighter until we reach point identification.

## Alternative Estimands

### Composite ATE (Ψ₁)

The composite ATE measures the effect on the composite outcome that combines observed outcomes with the event of missingness:

```{r}
fit_psi1 <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "psi1",
  assumption = "bounded_delta",
  delta_0u = 0.8,
  delta_1u = 0.8,
  sl_lib = "SL.glm",
  V = 2,
  seed = 1
)

fit_psi1$result
```

### Separable Direct Effect (Ψ₂)

The separable direct effect isolates the effect of treatment on the outcome among those who would have observed outcomes regardless of treatment:

```{r}
fit_psi2 <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "psi2",
  assumption = "point_psi2",
  delta_0 = 0.5,
  sl_lib = "SL.glm",
  V = 2,
  seed = 1
)

fit_psi2$result
```

## Smooth Approximation

For `bounded_delta` (with Ψ₂) and `bounded_risk` assumptions, you can use smooth approximation instead of indicator functions. This replaces discontinuous indicator functions with smooth CDF approximations:

```{r}
# Bounded delta with smooth approximation
fit_psi2_smooth <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "psi2",
  assumption = "bounded_delta",
  delta_0 = 0.5,
  sl_lib = "SL.glm",
  V = 2,
  seed = 1,
  smooth_approximation = TRUE,
  epsilon = 0.001  # Smoothing parameter
)

# Bounded risk with smooth approximation
fit_risk_smooth <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "ate",
  assumption = "bounded_risk",
  delta_0u = 0.8,
  delta_1u = 0.8,
  tau_0 = 2,
  tau_1 = 2,
  sl_lib = "SL.glm",
  V = 2,
  seed = 1,
  smooth_approximation = TRUE,
  epsilon = 0.001
)
```

Smooth approximations:
- Closely match indicator methods when `epsilon` is small (0.001-0.01)
- Provide differentiable bounds for optimization
- Are useful when discontinuous functions are problematic



## Customizing SuperLearner

You can specify different libraries for different nuisance functions:

```{r, eval = FALSE}
fit_custom <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "ate",
  assumption = "general",
  sl_lib_prop = c("SL.glm", "SL.mean"),      # Propensity score
  sl_lib_miss = c("SL.glm", "SL.gam"),       # Missingness probabilities
  sl_lib_outcome = c("SL.glm", "SL.ranger"), # Outcome regression
  V = 2,
  seed = 1
)
```

## Next Steps

- See `vignette("sensitivity-analysis")` for sensitivity analysis across parameter grids
- See `vignette("advanced")` for advanced options and technical details
- See the package documentation for details on each assumption type

## References

Rubinstein, M., Agniel, D., Han, L., Horvitz-Lennon, M., & Normand, S.-L. (2026).
Bounding causal effects with an unknown mixture of informative and non-informative missingness.
Journal of Causal Inference (Accepted). https://arxiv.org/pdf/2411.16902
