---
title: "Sensitivity Analysis with Parameter Grids"
author: "Max Rubinstein, Denis Agniel, Larry Han, Marcela Horvitz-Lennon, Sharon-Lise Normand"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity Analysis with Parameter Grids}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

When the sensitivity parameters (δ, τ) are unknown, we can perform sensitivity analysis by computing bounds across a grid of plausible parameter values. The **jointbounds** package provides simultaneous inference via multiplier bootstrap to control familywise error rate across the entire grid.

```{r}
library(jointbounds)
suppressPackageStartupMessages(library(SuperLearner))
```

## Example Data

Let's use a larger dataset for this example:

```{r}
set.seed(20260313)
n <- 1000

X <- runif(n, -2, 2)
A <- rbinom(n, 1, plogis(X))
C <- rbinom(n, 1, 0.2 + 0.1 * A)
Y <- ifelse(C == 1, NA, rbinom(n, 1, plogis(0.5 * A + 0.3 * X)))

dat <- data.frame(Y = Y, A = A, C = C, X = X)
```

## Grid Specification

### Basic Grid: Varying δ

The most common sensitivity analysis varies the proportion of informative missingness (δ):

```{r}
fit_delta_grid <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "ate",
  assumption = "bounded_delta",
  param_grid = list(
    delta_0u = seq(0.3, 1.0, by = 0.1),
    delta_1u = seq(0.3, 1.0, by = 0.1)
  ),
  sl_lib = "SL.glm",
  V = 2,
  alpha = 0.05,
  B = 500,  # Number of bootstrap replicates
  seed = 1
)
```

### Inspecting Grid Results

The output contains separate grids for lower and upper bounds:

```{r}
# Lower bound grid
head(fit_delta_grid$lower_grid$grid)

# Upper bound grid
head(fit_delta_grid$upper_grid$grid)
```

Each row in the grid contains:
- Parameter values (delta_0, delta_1)
- `estimate`: The bound estimate
- `se`: Standard error (same for pointwise and uniform inference)
- `ci_*_pointwise`: Pointwise confidence intervals
- `ci_*_uniform`: Simultaneous confidence intervals

### Critical Values

The multiplier bootstrap provides a critical value for simultaneous inference:

```{r}
cat("Lower bound critical value:", fit_delta_grid$lower_grid$critical_value, "\n")
cat("Upper bound critical value:", fit_delta_grid$upper_grid$critical_value, "\n")
cat("Standard normal quantile:", qnorm(0.975), "\n")
```

The critical value is larger than the standard normal quantile, reflecting the adjustment for multiple testing.

## Visualizing Sensitivity Analysis

### Lower Bound Across δ Values

```{r}
library(ggplot2)

grid <- fit_delta_grid$lower_grid$grid

# When both deltas are equal
diag_subset <- grid[grid$delta_0u == grid$delta_1u, ]

ggplot(diag_subset, aes(x = delta_0u)) +
  geom_line(aes(y = estimate), linewidth = 1) +
  geom_ribbon(aes(ymin = ci_lower_pointwise, ymax = ci_upper_pointwise),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(aes(ymin = ci_lower_uniform, ymax = ci_upper_uniform),
              alpha = 0.2, fill = "red") +
  labs(
    title = "Lower Bound Sensitivity Analysis",
    subtitle = "Blue = pointwise CIs, Red = simultaneous CIs",
    x = "Proportion of informative missingness (δ)",
    y = "Lower bound on ATE"
  ) +
  theme_minimal()
```

### Heatmap: Full 2D Grid

```{r}
ggplot(grid, aes(x = delta_0u, y = delta_1u, fill = estimate)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Lower\nBound") +
  labs(
    title = "Lower Bound Heatmap",
    x = "δ₀ (control group)",
    y = "δ₁ (treatment group)"
  ) +
  theme_minimal() +
  coord_fixed()
```

### Upper Bound

```{r}
upper_grid <- fit_delta_grid$upper_grid$grid
upper_diag <- upper_grid[upper_grid$delta_0u == upper_grid$delta_1u, ]

ggplot(upper_diag, aes(x = delta_0u)) +
  geom_line(aes(y = estimate), linewidth = 1) +
  geom_ribbon(aes(ymin = ci_lower_pointwise, ymax = ci_upper_pointwise),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(aes(ymin = ci_lower_uniform, ymax = ci_upper_uniform),
              alpha = 0.2, fill = "red") +
  labs(
    title = "Upper Bound Sensitivity Analysis",
    subtitle = "Blue = pointwise CIs, Red = simultaneous CIs",
    x = "Proportion of informative missingness (δ)",
    y = "Upper bound on ATE"
  ) +
  theme_minimal()
```

## Advanced Grid Options

### Bounded Risk Assumption

When using bounded risk assumptions, we can vary both δ and τ parameters:

```{r, eval = FALSE}
fit_risk_grid <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "ate",
  assumption = "bounded_risk",
  param_grid = list(
    delta_0u = seq(0.5, 1.0, by = 0.1),
    delta_1u = seq(0.5, 1.0, by = 0.1),
    tau_0 = c(1.5, 2.0, 2.5),
    tau_1 = c(1.5, 2.0, 2.5)
  ),
  sl_lib = "SL.glm",
  V = 2,
  B = 500,
  seed = 1
)

# This creates a 6 x 6 x 3 x 3 = 324 point grid
nrow(fit_risk_grid$lower_grid$grid)
```

### Controlling Which Bounds to Compute

For large grids, you may want to compute only one bound:

```{r}
# Only lower bound
fit_lower_only <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "ate",
  assumption = "bounded_delta",
  param_grid = list(
    delta_0u = seq(0.5, 1.0, by = 0.1),
    delta_1u = seq(0.5, 1.0, by = 0.1)
  ),
  grid_bounds = "lower",  # Only compute lower bound
  sl_lib = "SL.glm",
  V = 2,
  B = 500,
  seed = 1
)

names(fit_lower_only)
```

## Point Estimate Grids (Ψ₂)

For the separable direct effect (Ψ₂), point identification is possible. We can create grids of point estimates:

```{r}
fit_psi2_grid <- mar_bounds(
  dat,
  Y = "Y",
  A = "A",
  C = "C",
  X = "X",
  estimand = "psi2",
  assumption = "point_psi2",
  param_grid = list(
    delta_0 = seq(0.3, 0.9, by = 0.1)
  ),
  sl_lib = "SL.glm",
  V = 2,
  B = 500,
  seed = 1
)

# View point estimate grid
print(fit_psi2_grid$result)
```

Visualize:

```{r}
est_grid <- fit_psi2_grid$result

ggplot(est_grid, aes(x = delta_0)) +
  geom_line(aes(y = estimate), linewidth = 1) +
  geom_ribbon(aes(ymin = ci_lower_pointwise, ymax = ci_upper_pointwise),
              alpha = 0.2, fill = "blue") +
  geom_ribbon(aes(ymin = ci_lower_uniform, ymax = ci_upper_uniform),
              alpha = 0.2, fill = "red") +
  labs(
    title = "Ψ₂ Sensitivity Analysis",
    subtitle = "Blue = pointwise CIs, Red = simultaneous CIs",
    x = "δ₀",
    y = "Separable direct effect estimate"
  ) +
  theme_minimal()
```

## Interpreting Results

### Width of Confidence Bands

The uniform confidence intervals are wider than pointwise intervals, accounting for searching over the parameter grid:

```{r}
grid <- fit_delta_grid$lower_grid$grid
diag <- grid[grid$delta_0u == grid$delta_1u, ]

cat("Average pointwise CI width:",
    mean(diag$ci_upper_pointwise - diag$ci_lower_pointwise), "\n")
cat("Average uniform CI width:",
    mean(diag$ci_upper_uniform - diag$ci_lower_uniform), "\n")
cat("Ratio:",
    mean(diag$ci_upper_uniform - diag$ci_lower_uniform) /
    mean(diag$ci_upper_pointwise - diag$ci_lower_pointwise), "\n")
```

### Identifying Parameter Regions

Find the parameter values where the bound excludes zero:

```{r}
grid <- fit_delta_grid$lower_grid$grid

# Using pointwise CIs
significant_pointwise <- grid[grid$ci_lower_pointwise > 0, ]
cat("Pointwise: ", nrow(significant_pointwise), "out of", nrow(grid),
    "grid points exclude zero\n")

# Using uniform CIs
significant_uniform <- grid[grid$ci_lower_uniform > 0, ]
cat("Uniform: ", nrow(significant_uniform), "out of", nrow(grid),
    "grid points exclude zero\n")
```

## Computational Considerations

### Bootstrap Replicates

More bootstrap replicates give more accurate critical values but take longer:

```{r, eval = FALSE}
# Quick exploration: B = 100-200
fit_quick <- mar_bounds(..., B = 100)

# Final analysis: B = 500-1000
fit_final <- mar_bounds(..., B = 1000)
```

### Grid Size

Large grids can be time-consuming. Consider:

```{r, eval = FALSE}
# Coarse grid first
fit_coarse <- mar_bounds(
  ...,
  param_grid = list(
    delta_0 = seq(0.3, 1.0, by = 0.2),
    delta_1 = seq(0.3, 1.0, by = 0.2)
  )
)

# Fine grid in region of interest
fit_fine <- mar_bounds(
  ...,
  param_grid = list(
    delta_0 = seq(0.5, 0.8, by = 0.05),
    delta_1 = seq(0.5, 0.8, by = 0.05)
  )
)
```

## Summary

Sensitivity analysis via parameter grids allows you to:
1. Assess how bounds change across plausible parameter values
2. Identify regions where treatment effects can be detected
3. Provide simultaneous inference controlling familywise error rate
4. Visualize uncertainty in the presence of unverifiable assumptions

The multiplier bootstrap ensures that confidence statements are valid across the entire parameter grid, not just at individual points.

## References

Rubinstein, M., Agniel, D., Han, L., Horvitz-Lennon, M., & Normand, S.-L. (2026).
Bounding causal effects with an unknown mixture of informative and non-informative missingness.
Journal of Causal Inference (Accepted). https://arxiv.org/pdf/2411.16902
